Towards a Classification of Sporadic $j$-invariants

### 2023-04-12 TeX/draft.tex

\usepackage{amsmath,amsthm,amssymb}

@@

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}%[section]
\newtheorem*{theorem*}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem*{proposition*}{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem*{corollary*}{Corollary}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{algorithm}[theorem]{Algorithm}
\DeclareMathOperator{\gon}{gon}
%\numberwithin{equation}{section}

\renewcommand{\epsilon}{\varepsilon}
\renewcommand{\phi}{\varphi}
\renewcommand{\theta}{\vartheta}

\renewcommand\labelenumi{(\theenumi)}
\renewcommand{\theenumi}{\roman{enumi}}

@@

% Nongross real and imaginary parts
\renewcommand{\Im}{\operatorname{Im}}
\renewcommand{\Re}{\operatorname{Re}}
\newcommand{\Q}{\bQ}
\newcommand{\Z}{\bZ}
\newcommand{\PP}{\mathbb P}
\newcommand{\filip}[1]{{\color{cyan} \textsf{$\sun\sun\sun$ Filip: [#1]}}}
\newcommand{\davidnote}[1]{\marginpar{\footnotesize{\textcolor{Plum}{David: #1}}}}

@@

@@

\begin{algorithm}
\textbf{Input:} A non-CM $j$-invariant $j = j(E) \in \Q$.

\textbf{Output:} \textsc{False} if $j$ is not a sporadic $j$-invariant.

\begin{enumerate}
    \item Compute the adelic image $G$ of $E/\Q$ as a subgroup of $\GL_2(\hat{\Z})$ using Zywina's algorithm. Represent the output as some \emph{level} $M$ together with a subgroup $G$ of $\GL_2(\Z/M\Z)$.

    \item Reduce level.

    \item For all $12 < b \mid M$ compute the minimum degree of the field of definition of a point with respect to $G \mod b$, the image of $G$ in $\GL_2(\Z/b\Z)$. (For that, compute the orbits of $\langle G, -G \rangle \mod b$ on the elements of $(\Z/b\Z)^2$ of order $b$. Sort the orbits by size. Return ) 
    %\filip{We need to look at the orbits of $\langle G, -G \rangle \mod b$ on $(\Z/b\Z)^2$, not $\bP^1(\Z/b\Z)$. If we consider the action on $\bP^1(\Z/b\Z)$, the orbits might be too small and hence the degree we get. Furthermore, it is necessary to consider only the elements of $(\Z/b\Z)^2$ of order $b$.}
\end{enumerate}
\end{algorithm}

(In the \texttt{Magma} implementation, we have to be careful that \texttt{Magma} uses right actions, but in our mathematics, we use left action. We therefore use the transposes of the matrices in $\GL_2(\Z/M\Z)$.)

@@

Let $E/ \Q$ be an elliptic curve and $G\in \GL_2(\hat \Z)$ its adelic image. By $G(n)$ denote the image of its mod $n$ representation. Denote by $B_1(n)$ the subgroup of $\GL_2(\Z /n \Z)$ consisting of the upper triangular matrices with a $1$ in the upper left entry. Note that $X_{B_1(n)}=X_1(n)$. We say that a congruence group $\Gamma$ is of genus $g$ if $X_\Gamma$ is of genus $g$. We say that a point $P$ corresponds to an elliptic curve $E$ if $j(P)=j(E)$.

>>>>

\usepackage{amsmath,amsthm,amssymb}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}

\usepackage{colonequals}

@@

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}%[section]
\newtheorem*{theorem*}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem*{proposition*}{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem*{corollary*}{Corollary}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{question}[theorem]{Question}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
% \newtheorem{algorithm}[theorem]{Algorithm}
\DeclareMathOperator{\gon}{gon}
%\numberwithin{equation}{section}

\renewcommand{\epsilon}{\varepsilon}
\renewcommand{\phi}{\varphi}
\renewcommand{\theta}{\vartheta}
\newcommand{\Supp}{\operatorname{Supp}}
\renewcommand\labelenumi{(\theenumi)}
\renewcommand{\theenumi}{\roman{enumi}}

@@

% Nongross real and imaginary parts
\renewcommand{\Im}{\operatorname{Im}}
\renewcommand{\Re}{\operatorname{Re}}
\newcommand{\Q}{\bQ}
\newcommand{\Z}{\bZ}
\newcommand{\PP}{\mathbb P}
\newcommand{\filip}[1]{{\color{cyan} \textsf{$\sun\sun\sun$ Filip: [#1]}}}
\newcommand{\davidnote}[1]{\textcolor{Plum}{David: #1}}

\newcommand{\abbey}[1]{\textcolor{blue}{Abbey: #1}}

\newcommand{\sachi}[1]{\textcolor{purple}{Sachi: #1}}
% \newcommand{\travis}[1]{\textcolor{chartreuse}{Travis: #1}}

@@

\section{Background}

\begin{theorem}
Let $E/\Q$ be a non-CM elliptic curve, and let $\rho_E$ denote the adelic Galois representation of $E$. Let $M_E$ be any positive integer such that 
\[
\im \rho_E=\pi^{-1}(\im \rho_{E, M_E}).
\]
Then for any $x \in X_1(n)$ with $j(x)=j(E)$ we have $\deg(x)=\deg(f)\cdot \deg(f(x))$, where $f: X_1(n) \rightarrow X_1(\gcd(n,M_E))$ denotes the natural map. It follows that:
\begin{enumerate}
\item If $x$ is $\mathbb{P}^1$-isolated, then $f(x)$ is $\mathbb{P}^1$-isolated.
\item If $x$ is AV-isolated, then $f(x)$ is AV-isolated.
\item If $x$ is sporadic, then $f(x)$ is sporadic.
\end{enumerate}
\end{theorem}

\begin{proof}
This follows from the arguments \cite[$\S5.3$]{BELOV} and \cite[Theorem 4.3]{BELOV}. \abbey{This is mentioned in Remark 5.5 of \cite{BELOV} in particular. I'm not sure if we should include more details here.}
\end{proof}

@@

\begin{algorithm}
\KwIn{A non-CM $j$-invariant $j = j(E) \in \Q$.}
\KwOut{\textsc{False} if $j$ is not a sporadic $j$-invariant. \abbey{Update: Finite list of curves, degrees such that any isolated point on $X_1(n)$ associated to $E$ maps down under the natural projection map to a point in this list. If empty, $j$ is not isolated.)}}
Compute the adelic image $G$ of $E/\Q$ as a subgroup of $\GL_2(\hat{\Z})$ using Zywina's algorithm. Represent the output as some \emph{level} $N$ together with a subgroup $G$ of $\GL_2(\Z/N\Z)$\;
Reduce level\;
For all $12 < b \mid N$, using Algorithm~\ref{alg:compute_degree} compute the minimum degree of the field of definition of a point with respect to $G \mod b$, the image of $G$ in $\GL_2(\Z/b\Z)$\;
\Return{?} \;
    %\filip{We need to look at the orbits of $\langle G, -G \rangle \mod b$ on $(\Z/b\Z)^2$, not $\bP^1(\Z/b\Z)$. If we consider the action on $\bP^1(\Z/b\Z)$, the orbits might be too small and hence the degree we get. Furthermore, it is necessary to consider only the elements of $(\Z/b\Z)^2$ of order $b$.}
\end{algorithm}

(In the \texttt{Magma} implementation, we have to be careful that \texttt{Magma} uses right actions, but in our mathematics, we use left action. We therefore use the transposes of the matrices in $\GL_2(\Z/N\Z)$.)

\section{Reduce Level}

\abbey{Travis,Sachi, Abbey}

\begin{theorem}[\cite{BELOV}]\label{BELOVthm}
Let $E/\Q$ be a non-CM elliptic curve, and let $m$ be the product of 2, 3, and all primes $\ell$ where the $\ell$-adic Galois representation of $E$ is not surjective. Let $m_0$ be the level of the $m$-adic Galois representation associated to $E$ and let $f: X_1(n) \rightarrow X_1(\gcd(n,m_0))$ denote the natural map. If $x\in X_1(n)$ is an isolated point with $j(x)=j(E)$, then $f(x)\in X_1(\gcd(n,m_0))$ is an isolated point. 
\end{theorem}

Using Zywina's algorithm for the level of the adelic Galois representation as a starting point, we may obtain the level of the $m$-adic Galois representation associated to $E$ as in Theorem \ref{BELOVthm}.

\begin{algorithm}[H]
  \KwIn{$G\subseteq \GL_2(\Z/N\Z)$ such that $\im\rho_E=\pi^{-1}(G)$ and $N$ is the level}
  \KwOut{$G\subseteq \GL_2(\Z/m_0\Z)$ such that $G=\im\rho_{E,m_0}$ and $\im\rho_{E,m^{\infty}}=\pi^{-1}(\im\rho_{E,m_0})$ where $m$ is the product of the non-surjective primes and $2$ and $3$ (if they do not already divide $m$), and $m_0$ is minimal \abbey{Don't we need to include 2 and 3 in our definition of $m$? It is possible 2 and 3 may divide $m_0$ even if they are not surjective primes. If they are not necessary, then 2 and 3 need not divide $m_0$ in the final output.} } 
  \caption{Reduce level}%
  \label{alg:reduce_level}
  Let $N_E = \{\ell|N \text{ prime} \colon \rho_{E,\ell} \text{ is non-surjective} \}$\;
  Let $S_E= N_E \cup \{ 2, 3\}$\;
  Let $m_0 = \prod_{\ell \in S_E} \ell^{v_{\ell}(N)}$\;
  \For{$\ell \in \{ \ell | m_0 \colon \ell_0 \text{ prime} \}$}{
  \While{$\ell|m_0$ and $\#G/\#(G \pmod{m_0/p}) = p^4$}{
    $m_0 = m_0/p$\;
    $G = G \pmod{m_0}$\;
  }
  \If{not $p$ in $N_E$ and $\ord_p(m_0) = 1$ and $\#G/\#(G \pmod{m_0/p}) = \#\GL_2(\Z/p\Z)$}{
  $m_0 = m_0/p$\;
   $G = G \pmod{m_0}$\;}}
  \Return{$G,m_0$}
\end{algorithm}

  \begin{proposition}
    Algorithm~\ref{alg:reduce_level} is correct.
  \end{proposition}
  \begin{proof}
    Since $N$ is the level of the adelic representation and both $m$ and $m_0$ divides $N$, we must have that $\im\rho_{E, m^\infty} = \pi^{-1}(\im \rho_{E,m_0})$. This property is an invariant of the for-loop. The for-loop 
    terminates once as many powers of $\ell$ have been removed from $m_0$ 
    as possible while maintaining this invariant. Thus $m_0$ is the minimal such integer and is thus the level. 
  \end{proof}

\section{Compute Degree}

\noindent \textbf{Input:} $\mathcal{I}_1$ from step 1 \\
\textbf{Output:} For each $d \in \mathcal{I}_1$, compute degree(s) of closed point on $X_1(N)$ associated to $E$. Return ??? \abbey{update algorithm description}

\begin{algorithm}[H]\label{alg:compute_degree}
  \KwIn{$G\subseteq \GL_2(\Z/N\Z)$ such that $\im\rho_E=\pi^{-1}(G)$ and $N$ is the level; a bound $g \in \Z$}
  \KwOut{all degrees of $x$-coordinates of points that are $\geq g$} 
  Let $H \colonequals \langle G, -I_2\rangle \leq \GL_2(\Z/N\Z)$\;
  Compute the orbits of $H$ acting on $(\Z/N\Z)^2$\label{Inc-1}\;
  Filter the orbit representatives that have order equal to $N$\;
  Compute the orbit sizes ``degrees'' $s$ such that $s/2 \geq g$\label{DivBy2}\;
  Remove duplicates\label{RemDup}\;
  \Return{degrees}
\end{algorithm}

Note that only working with orbit(s) (representatives) significantly speeds up the implementation. The computation of the orbits of $G$ acting on $(\Z/N\Z)^2$ is efficiently implemented in Magma.

\begin{proposition}
Let $E/\Q$ be a non-CM elliptic curve and $N \in \Z_{>0}$. If $\im \rho_{E,N} \cong G \leq \GL_2(\Z/N\Z)$, then the Algorithm \ref{alg:compute_degree} outputs the sequence consisting of all possible degrees of $x(P)$, where $P\in E[N]$ is a point of order $N$.
\end{proposition}
\begin{proof}
    We begin by noting that $(E, P)$ and $(E, -P)$ denote the same point on $X_1(N)$, for $P \in E[N]$. Therefore, the degree of 
    a point represented by $(E, P)$, when
    $E$ is defined over $\Q$ depends only on $x(P)$. 
    If $E[N]$ is identified with $(\Z/N\Z)^2$, then $\deg(P)=\# (G. P)$, where $P$ on the left hand side is considered as an element of $(\Z/N\Z)^2$
    and on right hand side as an element of $E[N]$.
    If $P$ and $-P$ are in the same Galois orbit, i.e., $-I_2\in G$, then $[\Q(x(P)):\Q]=[\Q(P):\Q]/2$. On the other hand, one has $[\Q(x(P)):\Q]=[\Q(P):\Q]$, when 
    $-I_2\notin G$. 
    Therefore, computing the orbit sizes with respect to the 
    group $H\coloneqq \langle G, -I_2 \rangle$ (step~\ref{Inc-1} in Algorithm \ref{alg:compute_degree}) and dividing
    by $2$ (step~\ref{DivBy2}) handles both the cases, because in the later case (i.e., when $-I_2\notin G$) the orbit sizes will be doubled. 
    Finally, step~\ref{RemDup} makes sure that every degree occurs only once in the output sequence.
\end{proof}

\section{Riemann-Roch}

\noindent \textbf{Input:} Output from step 2\\
\textbf{Output:} Smaller list of pairs to check

\begin{lemma}
Let $C$ be a curve over a number field $k$ with genus $g$. If $x \in C(\overline{k})$ is a closed point with degree
$\geq g+1$, then $x$ is not $\mathbb{P}^{1}$-isolated.
\end{lemma}

\begin{proof}
This is Lemma 13 in revision of B., Gill, Rouse, Watson.
\end{proof}

\section{Remaining Cases}
\abbey{Filip? Abbey?}
Filip's arguments.

@@

Let $E/ \Q$ be an elliptic curve and $G\in \GL_2(\hat \Z)$ its adelic image. By $G(n)$ denote the image of its mod $n$ representation. Denote by $B_1(n)$ the subgroup of $\GL_2(\Z/n\Z)$ consisting of the upper triangular matrices with a $1$ in the upper left entry. Note that $X_{B_1(n)}=X_1(n)$. We say that a congruence group $\Gamma$ is of genus $g$ if $X_\Gamma$ is of genus $g$. We say that a point $P$ corresponds to an elliptic curve $E$ if $j(P)=j(E)$.
